---
title: "Primary Analysis: Cheatgrass and Wildlife"
author: "Anyll Markevich"
format: html
editor: visual
---

# Primary Analysis: Cheatgrass and Wildlife

## Objective

MENTION RESEARCH PAPER! This document details the analysis steps used to test wether observational data is consistent with the hypothesis that cheatgrass negatively impacts large mammals, birds, butterflies. More specifically, we test whether areas with higher cheatgrass cover correlate with lower abundance and species richness of the these three taxa. Although this is an observational study that cannot confidently draw causational conclusions, this analysis will compare the statistical explanatory power of cheatgrass cover against multiple other potentially relevant environmental factors to determine which factors are most likely to be causational. We will use Bayesian Model Averaging to achieve this objective.

## Setup

### Libraries

```{r}
#| output: false
library(here)
library(dplyr)
library(car)
library(ggplot2)
library(BMS)
library(performance)
library(patchwork)
```

```{r}
set.seed(60)
```

### Data Processing

The first step is to process raw data (found in `/data/raw/`) into analysis ready formats (saved to `/data/processed/`), all following tidy data conventions.

```{r}
#| output: false
source(here::here("R", "01_vegetation_survey_processing.R"))
source(here::here("R", "02_bird_count_processing.R"))
source(here::here("R", "03_butterfly_count_processing.R"))
source(here::here("R", "04_mammal_count_processing.R"))
source(here::here("R", "05_additional_data_processing.R"))
```

This tidy data can then be converted into summary tables that formatted for the following analysis (saved to `/data/summarized/`).

```{r}
#| output: false
source(here::here("R", "06_summarize_for_analysis.R"))
```

### Import Data

The analysis will require the summarized data for the primary analysis and a small data table that contains metadata about each column. This metadata will allow easy selection of just predictor our response variables, for instance, allowing for automation of repeated analyses.

```{r}
#| output: false
original_data <- read.csv(here::here(
  "data", 
  "summarized", 
  "main_summary.csv"))
original_data_columns <- read.csv(here::here(
  "data", 
  "summarized", 
  "main_summary_columns.csv"))
```

Note that this summary data only includes 4 of the 6 butterfly surveys conducted in 2020 to keep the methodology consistent with the 4 total butterfly surveys conducted in 2021.

### Calculate Metrics

Currently, cheatgrass cover is expressed as two separate values: cheatgrass plant cover and cheatgrass litter cover. Although distinguishing between these two types of cheatgrass prevalence could provide valuable ecological insight, this analysis aims to capture the overall impact of cheatgrass on wildlife. Thus, we will calculate and include a total cheatgrass cover value by summing the two cheatgrass metrics.

```{r}
data <- original_data
data$cheatgrass_total <- data$cheatgrass_plant + data$cheatgrass_litter
```

Lastly, I will update the metadata table to include these new data columns.

```{r}
data_columns <- original_data_columns
data_columns[nrow(data_columns) + 1, ] <- c("cheatgrass_total", "predictor")
```

### Avoiding Regression Singularity

To avoid singularity in the vegetation data, we will only include metrics directly relevant to cheatgrass and bare ground, removing the measurments of non-cheatgrass plant cover and non-cheatgrass litter cover. An additional singularity remains because `cheatgrass_total = cheatgrass_plan + cheatgrass_litter`, but this will be addressed later alongside for why certain predictors were kept in the final analysis.

```{r}
data <- data %>% 
  dplyr::select(- other_litter, - other_plant)
data_columns <- data_columns %>% subset(!(column %in% c("other_litter", "other_plant")))
```

### Prepare Data

SCALE DATA???

### Create Labels

```{r}
data_columns$label <- NA
data_columns$label[which(data_columns$type %in% c("response", "predictor", "category"))] <- 
  c("Mammal Count", "Mammal Species Richness", "Bird Count", "Bird Species Richness", "Butterfly Count", "Butterfly Species Richness", "Cheatgrass Plant Cover", "Cheatgrass Litter Cover", "Bare Ground Cover", "Woody Cover", "Living Woody Cover", "Distance to Highway", "Slope", "Year", "Plot in Burn Area", "Plot was Burned", "Cheatgrass Cover")
data_columns$predictor_order[which(data_columns$type %in% c("predictor", "category"))] <- 
  c(2, 3, 7, 5, 6 , 9, 8, 10, 11, 12, 1)
```

## Preparing Analysis

### Data Transformations

All the response variables are counts and therefore theoretically follow a Poisson distribution instead of a Gaussian distribution. Although it would be possible to run Bayesian Model Averaging with a Poisson-based GLM using the BMA package, we had significant overfitting issues when using this package. Meanwhile, the BMS package, which does not support Poisson distributions, provided enough control over priors for us to prevent large models from having high influence over the outcomes. Thus, we will use an Anscombe transform on response variables to approximate a gaussian distribution from count data. Zero inflation is not an issue in any of these data. This is to stablaize variance.

```{r}
non_gaussian_column_ids <- which(
    colnames(data) %in% subset(data_columns, type == "response")$column
  )
data[, non_gaussian_column_ids] <- 2*sqrt(data[, non_gaussian_column_ids] + (3/8))
```

### Addressing Multicollinearity

We will select of set of predictor variables for the analysis that are not strongly correlated with one another — predictor variables with low multicollinearity. Although Bayesian Model Averaging rigorously handles multicollinearity, the results may be more difficult to interpret as statistical power is split across multiple variables. We will use a pearson's correlation coefficient matrix to list all pairs of predictors that have a pearson's correlation coefficient over 0.8.

```{r}
temp_data <- data %>% 
  dplyr::select(
    data_columns$column[which(data_columns$type == "predictor")],
    data_columns$column[which(data_columns$type == "category")]
    )
corr_matrix <- cor(temp_data)
corr_matrix[upper.tri(corr_matrix, diag = TRUE)] <- NA
indices <- which(corr_matrix >= 0.8, arr.ind=TRUE)
cat(
  paste("Pairs of preditors with a correlation coefficient greater than or equal to 0.8:",
  paste(
    rownames(corr_matrix)[indices[,1]], 
    colnames(corr_matrix)[indices[,2]], sep = " and ", collapse = ",\n"
    ), 
  sep = "\n")
  )
```

All the cheatgrass cover variables are strongly correlated with one another (and form a regression singularity), so we will only keep the total cheatgrass cover metric (`cheatgrass_total`). The strong correlation between cheatgrass plants and litter suggest that this study is not capable of distinguishing between the ecological impact of these types of ground cover. Thus, we opt to use the sum of these two values as this combined metric is more likely to show a statistical relationship with the abundance and species richness of any taxa impacted by either or both cheatgrass plants and cheatgrass litter.

Similarly, as plot slope and the distance between the plot and the nearest highway are highly correlated due to the mountain geography in the study area, we will only keep the slope predictor (`slope`). Distance to the highway was originally included in case mammals were afraid of human activity, but we had no evidence that this was the case, so we will keep slope as it is a more ecologically relevant predictor.

```{r}
data <- data %>%
  dplyr::select(- cheatgrass_litter, - cheatgrass_plant, - distance_to_highway)
data_columns <- data_columns %>% subset(!(column %in% c("cheatgrass_litter", "cheatgrass_plant", "distance_to_highway")))
```

We also exclude the living woody cover predictor variable because it all values are near duplicates of the total woody cover except on the four burned plots post-burn (in 2021). Although living shrubs and trees likely play a different ecological role than their leafless counterparts, both provide shelter and habitat to numerous species.

```{r}
data <- data %>%
  dplyr::select(- living_woody_cover)
data_columns <- data_columns %>% subset(!(column %in% c("living_woody_cover")))
```

To ensure the multicollinearity won't be an issue, we will run a second test to ensure that all Variance Inflation Factor values are below 5.

```{r}
temp_data <- data %>% 
  dplyr::select(
    data_columns$column[which(data_columns$type == "predictor")],
    data_columns$column[which(data_columns$type == "category")]
    )
temp_data$example_response <- 1:nrow(data) # Create a dummy response variable
mod <- lm(example_response ~ ., data = temp_data)
vif_values <- car::vif(mod)
cat(paste("Predictors with VIF values greater than 5:", 
          names(vif_values[which(vif_values > 5)]) 
          ))
```

To address this issue, we exclude the bare ground predictor.

```{r}
data <- data %>%
  dplyr::select(- bare_ground)
data_columns <- data_columns %>% subset(!(column == "bare_ground"))
```

Most tests for linearity require more degrees of freedom to provide reliable results than can be spared in our small dataset. Thus, we take a visual estimation approach, checking partial regression plots for any obvious nonlinearity. We simultaneously test model diagnostics using the performance package.

```{r}
#| fig-height: 8
for (response in data_columns$column[which(data_columns$type == "response")]) {
  temp_data <- data %>%
    dplyr::select(
      all_of(response), 
      all_of(data_columns$column[which(data_columns$type == "predictor")]),
      all_of(data_columns$column[which(data_columns$type == "category")])
      )
  mod <- lm(temp_data)
  car::avPlots(mod, main = paste(
    "Partial Regression for", 
    data_columns$label[which(data_columns$column == response)]))
}
```

```{r}
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10
for (response in data_columns$column[which(data_columns$type == "response")]) {
  cat(paste("Diagnostic Plots for", response))
  temp_data <- data %>%
    dplyr::select(
      all_of(response), 
      all_of(data_columns$column[which(data_columns$type == "predictor")]),
      all_of(data_columns$column[which(data_columns$type == "category")])
      )
  mod <- lm(temp_data)
  plot(performance::check_model(mod))
}
```

Although the diagnostic plots are not perfect, variation and imperfections are acceptable given our small sample size. The only metric we found concerning was a single influential datapoint in the butterfly species richness regression with barely failed the cook's distance test (????). We considered removing this datapoint, but it caused a cascade of further outliers and Variance Inflation issues. Thus, we opted to leave this datapoint despite its high influence.

## Analysis

```{r}
#| results: false
bms_mods <- list()
for (response in data_columns$column[which(data_columns$type == "response")]) {
  analysis_data <- data %>% 
    dplyr::select(
      all_of(response), 
      all_of(data_columns$column[which(data_columns$type == "predictor")]), 
      all_of(data_columns$column[which(data_columns$type == "category")])
      )
  bms_mods[[length(bms_mods) + 1]] <- BMS::bms(
    analysis_data, 
    mprior = "fixed", 
    mprior.size = 2, 
    mcmc = "enumerate", 
    g = "hyper", 
    user.int = FALSE
    )
}
names(bms_mods) <- data_columns$column[which(data_columns$type == "response")]
```

## Results

```{r}
#| results: 'asis'
#| echo: false
for (response in data_columns$column[which(data_columns$type == "response")]) {
  cat(paste(
    "\n\n### Results for", 
    data_columns$label[which(data_columns$column == response)], 
    "Response\n\n```\n"
    ))
  bms_mod <- bms_mods[[response]]
  print(bms_mod)
  cat("```\n")
}
```

```{r}
#| fig-height: 8
#| fig-width: 8
bms_plots <- list()
for (i in 1:length(bms_mods)) {
  bms_mod <- bms_mods[[i]]
  effects <- data.frame(coef(bms_mod, std.coefs = TRUE))
  plot_data <- effects[, 1:3]
  plot_data$name <- rownames(effects)
  plot_data$label <- plot_data$name %>% vapply(function(x) {data_columns$label[which(data_columns$column == x)]}, FUN.VALUE = c(""))
  plot_data$order <- plot_data$name %>% vapply(function(x) {data_columns$predictor_order[which(data_columns$column == x)]}, FUN.VALUE = c(0))
  for (variable in plot_data$name) {
    confidence_intervals <- quantile(density(bms_mod, reg = variable, plot = FALSE, std.coefs = TRUE), probs = c(0.025, 0.975))
    plot_data$lower_ci[which(plot_data$name == variable)] <- confidence_intervals[1]
    plot_data$upper_ci[which(plot_data$name == variable)] <- confidence_intervals[2]
  }
  plot_data <- plot_data[order(plot_data$order),]
  plot_data$sequence <- 1:nrow(plot_data)
  forest_plot <- ggplot(plot_data, aes(x = Post.Mean, y = reorder(label, -sequence), xmin = lower_ci, xmax = upper_ci)) +
    geom_vline(xintercept=0, lty=2, color = "gray60") +
    geom_linerange() +
    geom_point(aes(size = PIP)) +
    scale_size_binned(breaks = c(0.3, 0.5, 0.7), limits = c(0, 1), range = c(0.5, 9)) +
    scale_x_continuous(breaks = c(-1, -0.5, 0, 0.5, 1), limits = c(-1.25, 1.25)) +
    theme_bw() + labs(
      x = "Standardized Posterior Mean (± 95% Credible Interval)",
      y = "Predictor",
      title = data_columns$label[which(data_columns$column ==names(bms_mods)[i])]
      ) + 
    theme(title =element_text(size = 8, face = 'plain'))
  bms_plots[[i]] <- forest_plot
}
patchwork::wrap_plots(bms_plots, ncol = 2) + 
  plot_layout(guides = "collect", axes = "collect") +
  plot_annotation(title = "Bayesian Model Averaging Results Summary") & 
  theme(
    axis.title = element_text(size = 12, face = "plain"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11, face = "plain")
    )
```
