---
title: "Secondary Analysis: Wildfire and Cheatgrass"
author: "Anyll Markevich"
format: html
editor: visual
---

# Secondary Analysis: Cheatgrass and Wildlife

## Objective

## Setup

### Libraries

```{r}
#| output: false
library(here)
library(dplyr)
library(rstanarm)
library(ggplot2)
library(report)
library(performance)
```

```{r}
set.seed(60)
```

### Data Processing

The first step is to process raw data (found in `/data/raw/`) into analysis ready formats (saved to `/data/processed/`), all following tidy data conventions.

```{r}
#| output: false
source(here::here("R", "01_vegetation_survey_processing.R"))
source(here::here("R", "02_bird_count_processing.R"))
source(here::here("R", "03_butterfly_count_processing.R"))
source(here::here("R", "04_mammal_count_processing.R"))
source(here::here("R", "05_additional_data_processing.R"))
```

This tidy data can then be converted into summary tables that formatted for the following analysis (saved to `/data/summarized/`).

```{r}
#| output: false
source(here::here("R", "06_summarize_for_analysis.R"))
```

### Import Data

The analysis will require the summarized data for the primary analysis and a small data table that contains metadata about each column. This metadata will allow easy selection of just predictor our response variables, for instance, allowing for automation of repeated analyses.

```{r}
#| output: false
original_data <- read.csv(here::here(
  "data", 
  "summarized", 
  "vegetation_summary.csv"))
```

## Process Data

```{r}
data <- original_data
data$cheatgrass_2020 <- data$cheatgrass_plant_2020 + data$cheatgrass_litter_2020
data$cheatgrass_2021 <- data$cheatgrass_plant_2021 + data$cheatgrass_litter_2021
```

## Diagnostics

```{r}
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10
ls_mod <- lm(cheatgrass_2021 ~ burned * cheatgrass_2020, data = data)
plot(performance::check_model(ls_mod))
```

## Analysis

```{r}
#| output: false
# Default weakly informative prior
def_bay_cheat <- stan_glm(
  cheatgrass_2021 ~ burned * cheatgrass_2020, 
  data = data, 
  family = gaussian(),
  refresh = 0
  )
default_priors <- prior_summary(def_bay_cheat)$prior$adjusted_scale
default_intercept <- prior_summary(def_bay_cheat)$prior_intercept$adjusted_scale
# Informative Prior
mod <- stan_glm(
  cheatgrass_2021 ~ burned * cheatgrass_2020, 
  data = data, 
  family = gaussian(), 
  prior = normal(
    location = c(0,1,0), 
    scale = c(default_priors[1], 0.5, default_priors[3]), 
    autoscale = FALSE
    ), 
  prior_intercept = normal(0, default_intercept, autoscale = FALSE),
  refresh = 0
  )
```

## Bayesian Diagnostics

```{r}
graph_distribution <- function(rstanarm_object, parameter) {
  posterior <- as.data.frame(rstanarm_object)
  unform_mod <- update(rstanarm_object, prior = NULL, prior_intercept = NULL)
  likelihood <- as.data.frame(unform_mod)
  n_params <- length(colnames(posterior))
  selected_param <- which(colnames(posterior) == parameter)
  if (selected_param == 1) {
    
  } else if (selected_param == n_params) {
    
  } else {
    predictor = selected_param - 1
    info <- prior_summary(rstanarm_object)$prior
    if (info$dist == "normal") {
      dist <- "dnorm"
      args <- list(mean = info$location[predictor], sd = info$scale[predictor])
    }
  }
  p <- ggplot() + 
    geom_density(data = posterior, mapping = aes(x = .data[[parameter]], color = "Posterior", fill = "Posterior"), inherit.aes = FALSE) + 
    stat_function(fun = eval(parse(text=dist)), args = args, mapping = aes(color = "Prior", fill = "Prior"), geom = "area") + 
    geom_density(data = likelihood, mapping = aes(x = .data[[parameter]], color = "Likelihood", fill = "Likelihood"), inherit.aes = FALSE) + 
    theme_classic() + 
    scale_fill_manual(values = c("Posterior" = "#7eb0d550", "Likelihood" = "#fd7f6f50", "Prior" = "#b2e06150"), name = "Distribution") +
    scale_color_manual(values = c("Posterior" = "#7eb0d5", "Likelihood" = "#fd7f6f", "Prior" = "#b2e061"), name = "Distribution") + 
    labs(title = "Bayesian Distributions for Diagnostics", y = "Probability Density")
  p
}

graph_distribution(mod, "burned")
graph_distribution(mod, "cheatgrass_2020")
graph_distribution(mod, "burned:cheatgrass_2020")
```

## Results

```{r}

```
